Stage-Splitting a Modal Language

### 2014-10-08 paper/relatedwork.tex
wrote stuff in related work

\subsection{Multi-stage (meta) programming}
{\bf Umut: jotting down notes for now.}

@@

\subsection{Specialization}

Our stage-splitting algorithm was first suggested in \cite{jorring86} under the name {\em pass separation}.  They essentially proposed that a function $f$ could be split into two others, $f_1$ and $f_2$, such that $f(x,y)=f_2(f_1(x),y)$.  They did not distinguish binding time analysis from stage splitting, and so pass separation inherits the former's ambiguity.  The main goal of \cite{jorring86} was to motivate pass separation and other staging transformations as a powerful way to think about compilation and optimization.  Accordingly, their approach was entirely informal, with no implementation realized.  Moreover, they predicted that ``the [pass separation] approach will elude full automation for some time."

Implementations of the stage-splitting algorithm have appeared in the literature exclusively (and coincidentally) in the context of graphics pipelines.  The first of these (\cite{knoblock96}) uses a binding time analysis to separate those parts of graphics shaders that are input-invariant from those which are not, and then uses a stage splitting algorithm to factor that into two shaders, thereby minimizing recomputation.  Their shaders are written in a C-like language with basic arithmetic and if statements.  Although their analysis does not give an explicit account of the type-level behavior of the splitting algorithm, it effectively can synthesize product and sum boundary type.

\subsection{Graphics Systems}

Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.  But instead of using a binding-time analysis, Spark allows the programmer to manually target stages of the graphics pipeline.  Since the modern graphics pipeline is inherently a many-to-one system, this is difficult to reconcile with sum types on the boundary.  Fortunately, Spark has a set of syntactic restrictions that prevent sum boundary types.  Spark does not clearly identify this conflict, but the authors did note that first-stage if statements were difficult to provide meaning to [need quote].

[RTSL and SH]

>>>>

\subsection {Staged Computation}
Partial evaluation is a well-studied technique for specializing programs to some of their input.  
The usual phrasing of the partial evaluation problem implicitly includes binding time analysis,
but many presentations start with a stage-annotated representation, as we have done.
We've shown that our dynamic semantics can be used to perform partial evaluation 
[need to do more reading to know exactly how it compares].

One unifying feature of partial evaluation systems is that they operate on code which is closed at the first stage,
and so they must necessarily unfold recursive calls.
This potentially explodes the size of a residual, which can be a problem.
Stage-splitting, in contrast, operates on terms which are open on first stage variables,
and so this problem is avoided.

@@

\subsection{Stage Splitting}

Stage-splitting itself has appeared in the literature---under 
the names {\em pass separation} and {\em data specialization}--- for various simpler input languages.

Most often, splitting is considered a compiler optimization (\cite{jorring86},\cite{knoblock96})
to minimize recomputation.
In this use case, only the arguments to the program are annotated with stages, 
and the splitter is left to perform an {\em ad hoc} binding time analysis to determine the 
internal staging of the program.
Because this occurs without programmer supervision, the binding time analyzers for these uses are 
reluctant to place stage \bbone\ code within stage \bbtwo\ conditionals, 
since that would result in speculative behavior that could increase the overall runtime.
We don't worry about this since the annotations of \lang\ are assumed to represent the will of a programmer.

Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.  
But instead of using a binding-time analysis to determine what parts of the program to put in what shader, 
Spark gives the programmer manual control of term staging using a language that is in spirit similar to \lang.

We are also not the first to use stage splitting to derive algorithms.
\cite{malmkjaer89}, expanding on the work of \cite{barzdins88}, defined a stage-splitter
for a language of tail-recursive first-order equations with explicit staging annotations [double check that],
and considered applications to string matching.

Our work is distinct from all previous stage splitters in the following ways:
\begin{itemize}
\item Our input language, \lang, has a disciplined type theory.
\item \lang\ has full first-class functions (and by extension, recursion), as well as sum types.
\item Our stage splitting is proven correct.
\end{itemize}
